---
title: "HRRP Heart Failure Readmissions"
author: "Nhi Nguyen"
date: "2025-12-22"
output: github_document
---
## Introduction

This report examines U.S. hospital performance on 30-day excess readmissions for heart failure under Medicare’s Hospital Readmissions Reduction Program (HRRP). The analysis uses hospital-level data for the heart failure measure READM-30-HF to study how excess readmission ratios vary across states and with hospital volume. Excess readmission ratios are risk-standardized performance metrics, where values above 1 indicate more readmissions than expected and values below 1 indicate better-than-expected outcomes.

The main empirical question is whether hospital volume and geography help explain variation in excess readmission performance for heart failure. To address this, the analysis combines SQL-based data cleaning with regression models in R, focusing on the log of the number of heart failure discharges and state fixed effects, and evaluates key modeling assumptions using formal diagnostics and robust standard errors.

## Research questions

The analysis is organized around three questions:

1. How does the excess readmission ratio for heart failure (READM-30-HF) vary across U.S. states?  
2. Are higher-volume hospitals, measured by the number of heart failure discharges, associated with better or worse excess readmission ratios after accounting for state differences?  
3. How much additional variation in excess readmission ratios is explained by state-level differences, beyond what can be captured by hospital volume alone?

```{r}
#Upload data
# Load the cleaned CMS HRRP heart failure data set exported from PostgreSQL
data <- read.csv("../data/cleaned_hrrp.csv")

# Quick structure and sanity check
# Confirm variable types and inspect the first few rows before analysis.
str(data)
head(data)

```
## Data and methods

The analysis uses hospital-level data from the CMS HRRP heart failure measure READM-30-HF. Each row represents a single hospital’s risk-standardized 30-day excess readmission performance for heart failure.

The main outcome is the excess readmission ratio, where values above 1 indicate higher-than-expected readmissions and values below 1 indicate better-than-expected performance. Hospital volume is measured using the number of heart failure discharges, and the models use the log of this variable to reduce skewness and obtain a more linear relationship. To capture geographic differences, the regressions include state fixed effects. Ordinary least squares (OLS) is used for estimation, with additional checks for heteroskedasticity and heteroskedasticity-consistent (HC1) standard errors for robust inference.

The data cover the measurement period from July 1, 2020 through June 30, 2023. After removing records with missing values for discharge counts and readmission ratios, the final sample includes 2,342 hospitals across 51 jurisdictions (50 states plus Washington D.C.). The number of heart failure discharges per hospital ranged from 30 to 3,490, with a median of 276.5. The excess readmission ratio ranged from 0.89 to 1.07, with 53.20% of hospitals performing worse than expected (ratio > 1).
```{r}
pct_worse <- sum(data$Excess.Readmission.Ratio > 1) / nrow(data) * 100
print (pct_worse)
```

## Exploratory data analysis
### Distribution of excess readmission ratios
Examine how excess readmission ratio is distributed to understand its center, spread, and skew.
```{r}
#EDA
library(ggplot2)
library(dplyr)

excess_ratio <- data$Excess.Readmission.Ratio
ggplot(data, aes(x = Excess.Readmission.Ratio)) +
  geom_histogram(color = "black", fill = "lightpink") +
  labs(title = "Distribution of Excess Readmission Ratios",
       x = "Excess Readmission Ratio", y = "Count") +
  theme_minimal()
```

### Distribution of raw discharge volume
Look at the skewness of hospital volume on the original scale.
```{r}
ggplot(data, aes(x = Number.of.Discharges)) +
  geom_histogram(color = "black", fill = "lightblue") +
  labs(title = "Distribution of Discharges",
       x = "Number of Discharges", y = "Count") +
  theme_minimal()
```

### Distribution of log(discharge volume)
Log-transform volume to compress the right tail and make patterns easier to see.
```{r}
ggplot(data, aes(x = log(Number.of.Discharges))) +
  geom_histogram(color = "black", fill = "lightblue") +
  labs(title = "Distribution of log(Discharges)",
       x = "log(Number of Discharges)", y = "Count") +
  theme_minimal()
```

### Geographic patterns
```{r}
# Define regions and explore variation across states
# Create a broad U.S. region variable (Northeast, Midwest, South, West)

data <- data %>%
  mutate(region = case_when(
    state %in% c("ME","NH","VT","MA","RI","CT","NY","NJ","PA") ~ "Northeast",
    state %in% c("OH","IN","IL","MI","WI","MN","IA","MO","ND","SD","NE","KS") ~ "Midwest",
    state %in% c("DE","MD","DC","VA","WV","NC","SC","GA","FL","AL","MS","TN","KY","AR","LA","OK","TX") ~ "South",
    state %in% c("MT","ID","WY","CO","UT","NV","CA","OR","WA","AK","HI","AZ","NM") ~ "West",
    TRUE ~ "Other"
  ))

# Boxplots by region: visualize how excess readmission ratios differ across states within each region.
```

West Region
```{r}
data_west <- subset(data, region == "West")

ggplot(data_west, aes(x = state, y = Excess.Readmission.Ratio)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Excess Readmission Ratios - West Region",
       x = "State", y = "Excess Readmission Ratio") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45))
```

Midwest Region
```{r}
data_midwest <- subset(data, region == "Midwest")
ggplot(data_midwest, aes(x = state, y = Excess.Readmission.Ratio)) +
  geom_boxplot(fill = "lightpink") +
  labs(title = "Excess Readmission Ratios - Midwest Region",
       x = "State", y = "Excess Readmission Ratio") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45))
```

South Region
```{r}
data_south <- subset(data, region == "South")
ggplot(data_south, aes(x = state, y = Excess.Readmission.Ratio)) +
  geom_boxplot(fill = 'lightgreen') +
  labs(title = "Excess Readmission Ratios - South Region",
       x = "State", y = "Excess Readmission Ratio") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45))
```

Northeast Region
```{r}
data_northeast <- subset(data, region == "Northeast")
ggplot(data_northeast, aes(x = state, y = Excess.Readmission.Ratio)) +
  geom_boxplot(fill = "violet") +
  labs(title = "Excess Readmission Ratios - Northeast Region",
       x = "State", y = "Excess Readmission Ratio") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45))
```

### Relationship between hospital volume and performance
Check how excess readmission ratios vary with hospital volume
```{r}
ggplot(data, aes(x = Number.of.Discharges, y = Excess.Readmission.Ratio)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Excess Readmission Ratio vs Discharges",
       x = "Number of Discharges",
       y = "Excess Readmission Ratio")
```

Use log(discharge volume) to see whether the association looks more linear
```{r}
ggplot(data, aes(x = log(Number.of.Discharges), y = Excess.Readmission.Ratio)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Excess Readmission Ratio vs Discharges",
       x = "Logarithms of Number of Discharges",
       y = "Excess Readmission Ratio")
```

## Regression specification

To quantify how hospital volume and geography are related to excess readmission performance, the report estimates three linear models using the excess readmission ratio as the dependent variable:

- **Model 1 (volume only)**  
  \[
  \text{ExcessRatio}_i = \beta_0 + \beta_1 \log(\text{Discharges}_i) + \varepsilon_i
  \]
  This specification asks how much variation in excess readmission ratios can be explained by hospital volume alone.

- **Model 2 (state only)**  
  \[
  \text{ExcessRatio}_i = \alpha_0 + \sum_s \alpha_s \mathbf{1}\{\text{state}_i = s\} + u_i
  \]
  This specification captures purely geographic differences across states without using volume.

- **Model 3 (volume + state, main model)**  
  \[
  \text{ExcessRatio}_i = \gamma_0 + \gamma_1 \log(\text{Discharges}_i) + \sum_s \gamma_s \mathbf{1}\{\text{state}_i = s\} + e_i
  \]
  This is the main specification and estimates the association between hospital volume and excess readmission ratios while controlling for state-level differences.

Comparing these models using \(R^2\) and a nested F-test shows how much additional variation is explained by adding state fixed effects to the volume-only model, and by including volume in a state-only model.

## Regression Results

### Model 1: volume-only model
```{r}
model1 <- lm(Excess.Readmission.Ratio ~ log(Number.of.Discharges), data = data)
summary(model1)
```
The volume-only model regresses the excess readmission ratio on the log of the number of heart failure discharges. In this specification, the coefficient on \(\log(\text{Discharges})\) is negative and statistically significant, indicating that higher-volume hospitals tend to have slightly lower excess readmission ratios on average. However, the model’s \(R^2\) is 0.0147 and the adjusted \(R^2\) is 0.0143, which are both very low. This implies that hospital volume alone explains only a small fraction of the cross-hospital variation in excess readmission performance.

### Model 2: state-only model
```{r}
model2 <- lm(Excess.Readmission.Ratio ~ factor(state), data = data)
summary(model2)
```
The state-only model includes state fixed effects but no volume term. In this case, the \(R^2\) is 0.126 and the adjusted \(R^2\) is 0.107, substantially higher than in Model 1. Even after penalizing for the larger number of parameters, the adjusted \(R^2\) remains much higher, indicating that state-level differences provide meaningful explanatory power beyond volume alone.

### Model 3: volume + state (main specification)
```{r}
model3 <- lm(Excess.Readmission.Ratio ~ log(Number.of.Discharges) + factor(state), data = data)
summary(model3)
```
The main model combines \(\log(\text{Discharges})\) with state fixed effects. Model 3 attains the highest \(R^2\) (0.150) and the adjusted \(R^2\) (0.131) of the three specifications. This suggests that combining hospital volume and state fixed effects yields the best overall fit, and that volume continues to add information even after accounting for geographic differences. 
The coefficient on \(\log(\text{Discharges})\) remains negative and statistically significant, meaning that, conditional on state, higher-volume hospitals have modestly lower excess readmission ratios on average.

```{r}
coef_log_discharge <- coef(model3)["log(Number.of.Discharges)"]

# Calculate effect of moving from 25th to 75th percentile
q25 <- quantile(log(data$Number.of.Discharges), 0.25)
q75 <- quantile(log(data$Number.of.Discharges), 0.75)
effect <- coef_log_discharge * (q75 - q25)

print(paste("Moving from 25th to 75th percentile in volume is associated with a", 
            round(effect, 3), "change in excess readmission ratio"))
```

## Diagnostics and Robust Inference
To check whether the homoscedastic assumption holds in the main model, the Breusch–Pagan test is applied to Model 3.

### Breusch-Pagan test for heteroskedasticity
```{r}
library(lmtest)
library(sandwich)

bptest(model3)
```
The test strongly rejects the null hypothesis of homoscedastic errors (p-value near zero), providing evidence of heteroskedasticity in the residuals. In this setting, the OLS coefficient estimates remain unbiased under the usual assumptions, but the conventional standard errors, t-statistics, and p-values are no longer reliable.

To address this, the analysis recomputes standard errors using heteroskedasticity-consistent (HC1) estimators via the `sandwich` package and reports robust inference using `lmtest::coeftest`. The robust standard errors are larger than the naive OLS standard errors, but the key coefficient on \(\log(\text{Discharges})\) remains statistically significant, reinforcing the conclusion that higher-volume hospitals tend to have slightly lower excess readmission ratios even after accounting for heteroskedasticity.

```{r}
coeftest(model3, vcov = vcovHC(model3, type="HC1"))
```

### Nested F-test for added state effects
```{r}
anova(model1,model3)
```
A nested F‑test comparing Model 1 (volume only) and Model 3 (volume + state) is used to assess the joint contribution of the state fixed effects. The F statistic is 7.31 with a p‑value < 0.001, showing that the state indicators as a group significantly improve model fit relative to a model with volume alone.

### Diagnostic plots for Model 3
Check linearity + equal variance
```{r}
plot(model3, which=1)
```

Check normality of residuals
```{r}
plot(model3, which=2)
```

Another view of homoscedasticity (spread of residuals)
```{r}
plot(model3, which=3)
```

Identify influential observations
```{r}
plot(model3, which=5)
```

Standard diagnostic plots for Model 3 support the basic linear specification but reveal non-constant error variance. The residuals-versus-fitted and scale–location plots show a changing spread of residuals across the fitted values, consistent with the Breusch–Pagan test and motivating the use of robust standard errors. The normal Q–Q plot shows some deviations in the tails but no extreme departures from normality that would make linear regression unusable in this context.

The residuals-versus-leverage plot highlights a small number of influential observations, but none appear to dominate the regression. Overall, a linear model in \(\log(\text{Discharges})\) with state fixed effects provides a reasonable summary of the relationship between hospital volume, geography, and excess readmission performance, as long as inference is based on heteroskedasticity-robust standard errors.

## Discussion

The results show that hospital volume and state-level factors are both associated with heart failure readmission performance, though state differences explain substantially more variation. The volume-only model explained less than 2% of variation in readmission ratios, while state fixed effects alone explained approximately 12%. The coefficient on log(discharges) is statistically significant in all models (β = -0.014, robust SE = 0.0017, p < 0.001), but the practical magnitude is modest: moving from the 25th to 75th percentile in discharge volume is associated with only a 0.017-point decrease in the excess readmission ratio.

State fixed effects explain a meaningful share of variation in excess readmission ratios, and nested F-tests confirm that state-level factors jointly add explanatory power beyond what can be captured by volume alone. This pattern suggests that broader policy, practice, or resource environments at the state level are important for understanding hospital readmission outcomes.At the same time, the presence of heteroskedasticity emphasizes the importance of using robust standard errors when interpreting regression results based on observational health care data.

There are several limitations in my project. First, this is a cross-sectional analysis and cannot support causal claims about the effect of volume on readmission outcomes. Second, while the HRRP measure is risk-adjusted for patient characteristics, I do not observe hospital-level factors such as nurse staffing ratios, teaching status, or participation in care coordination programs that may affect readmission performance. Third, the modest R² values (15% in the full model) indicate that most variation in readmission performance remains unexplained by volume and state alone. Future extensions could incorporate panel data across multiple years, add hospital-level covariates, or examine whether the volume-outcome relationship differs for safety-net or rural hospitals.

